---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# szKendall

<!-- badges: start -->
<!-- badges: end -->

The goal of szKendall is to measure the dissimilarity between two single cells based on their observed Hi-C data, accounting for the discrepancies in structural zero positions.

## Installation

You can install the szKendall package from GitHub:

``` r
library(devtools)
devtools::install_github("https://github.com/sl-lin/szKendall")
```

## Example

This is an example which shows you how to calculate the szKendall dissimilarity between cells. The simulated data contain 150 single cells (i.e., 150 columns). Therefore, the resulting szKendall matrix (which is symmetric with the main-diagonal all being 0) is of dimension 150-by-150. 

```{r example}

# From GitHub
# install.packages("devtools")
library(devtools)
install_github("https://github.com/sl-lin/szKendall")

library(szKendall)

# Example code
data("sim1.data")
data("true1.data")
# Calculate szKendall dissimilarity 
doParallel::registerDoParallel(cores = 2)  # Run registerDoSEQ() to override any default parallel backend and force sequential calculation 
# foreach::registerDoSEQ()
szK.sim1 <- szKendall.diss(sim1.data, true1.data)
szK1.sim1 <- szKendall1.diss(sim1.data, true1.data)
szK2.sim1 <- szKendall2.diss(sim1.data, true1.data)

print(dim(szK.sim1))
print(dim(szK1.sim1))
print(dim(szK2.sim1))
```


