---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# szKendall

<!-- badges: start -->
<!-- badges: end -->

The goal of szKendall is to measure the dissimilarity between two single cells based on their observed Hi-C data, accounting for the discrepancies in structural zero positions.

## System Requirements
- R version: >= 3.5
- Tested on: macOS 15.5, Windows 11
- Dependencies: None 
- No special hardware required


## Installation

You can install the szKendall package from GitHub:

```{r, eval = FALSE}
library(devtools)
devtools::install_github("https://github.com/swl0923/szKendall_trial")
```

The installation would take a few minutes on a standard desktop computer. 


## Example

This is an example which shows you how to calculate the szKendall dissimilarity between cells.The `PFC_subset` dataset consists of chr21 single-cell Hi-C contact matrices generated from 14 human prefrontal cortex (PFC) cell types, with 10 cells sampled from each type. Each matrix represents loci-pair-by-cell contact profiles that have been pre-processed using \pkg{HiCImpute}. With sequential calculation, the calculation time of one szKendall dissimilarity in this example should be less than 2 minutes, and half (or even less) with parallel computing. 

```{r example}
library(szKendall)
data("PFC_subset")

foreach::registerDoSEQ()
res <- szKendall(PFC_subset,dist.method="szKendall")
```

`szKendall()` returns a list of three \(n \times n\) symmetric matrices:

1. the **szKendall dissimilarity** matrix

2. the **Euclidean distance** matrix

3. the **Kendall dissimilarity** matrix

in that order.

```{r, fig.width=12, fig.height=4, dpi=300}
par(mfrow=c(1,3), mar = c(4,4,2,5), oma = c(0,0,0,2))
scHiC_hm_dist(res[[1]],title="szKendall")
scHiC_hm_dist(res[[2]],title="Euclidean")
scHiC_hm_dist(res[[3]],title="Kendall")
```



