---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# szKendall

<!-- badges: start -->
<!-- badges: end -->

The goal of szKendall is to measure the dissimilarity between two single cells based on their observed Hi-C data, accounting for the discrepancies in structural zero positions.

## System Requirements
- R version: >= 3.5
- Tested on: macOS 15.5, Windows 11
- Dependencies: None 
- No special hardware required


## Installation

You can install the szKendall package from GitHub:

```{r, eval = FALSE}
library(devtools)
devtools::install_github("https://github.com/swl0923/szKendall_trial")
```

The installation would take a few minutes on a standard desktop computer. 


## Example

This is an example which shows you how to calculate the szKendall dissimilarity between cells. The `PFC_subset_matrix` dataset consists of chr21 single-cell Hi-C contact matrices generated from 14 human prefrontal cortex (PFC) cell types, with 10 cells sampled from each type. Each matrix represents loci-pair-by-cell contact profiles that have been pre-processed using \pkg{HiCImpute}.
With sequential calculation, the calculation time of one szKendall dissimilarity in this example is less than 1 minute using 19 cores.
It may take longer when using fewer cores.

```{r example}
library(szKendall)
data("PFC_subset_matrix")

foreach::registerDoSEQ()
res <- szKendall(PFC_subset_matrix,dist.method="szKendall")
```


`szKendall()` takes three types of input formats:

- a list of scHi-C 2D matrices of size $n$ (e.g., number of bins per chromosome)
- a 3D array of dimension $(n \times n \times \text{number of cells})$
- a $(\text{number of locus-pairs} \times \text{number ofcells})$ matrix


```{r eval = FALSE}
# A list n*n matrices, one per cell. 
data("PFC_subset_list")
szKendall(PFC_subset_list,dist.method="szKendall")

# An array of dimension n * n * (number of cells)
data("PFC_subset_array")
szKendall(PFC_subset_array,dist.method="szKendall")
```


`szKendall()` returns a list of three \(n \times n\) symmetric matrices:

1. the **szKendall dissimilarity** matrix
2. the **Euclidean distance** matrix
3. the **Kendall dissimilarity** matrix

in that order.

```{r}
dim(res[[1]]) # szKendall or szKendall1
dim(res[[2]]) # Euclidean 
dim(res[[3]]) # Kendall 
```

<!-- ```{r, fig.width=12, fig.height=4, dpi=300}
par(mfrow=c(1,3), mar = c(4,4,2,5), oma = c(0,0,0,2))
scHiC_hm_dist(res[[1]],title="szKendall")
scHiC_hm_dist(res[[2]],title="Euclidean")
scHiC_hm_dist(res[[3]],title="Kendall")
``` -->



